They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-20 paper/paper.Rmd
more methodology writing

The formula for the expected proportion of seriesn a best-of-$s$ format ($s = 11$ rounds for CoD SnD), ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^11] is

@@

$$
m = \frac{s + 1}{2}
$$

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m).
}\label{eq:series-length}\end{equation}
```
To evaluate the constant round win probability hypothesis--that is, that the expected and observed round win rates, $\hat{\Phi}(r)$ and $\Phi(r)$ respectively, are equal to one another--we can compute the chi-square goodness of fit statistic, $\chi^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}
}\label{eq:chi-squ}\end{equation}
```
For CoD SnD, $r \in R = [6, 7, 8, 9, 10, 11]$.

@@

Removing our knowledge of a streak, we might model the Bernoulli round win probability as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwr}{}{
p^{w|r} = p(\text{win | r})=\{
\begin{array}{cl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}.
}
\label{eq:pwr}
\end{equation}
```
The Bernoulli round loss probability after a streak of $k$ round wins, $p^{\ell|kr}_0$, can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}_{MS}$, to be strictly less than $p^{w|r}$.[^12]

[^12]: See Appendix E.1 of MS for details on the proof that follows.

> Theorem 1: *Let* $\mathbf{X} = \{X_i\}^{n}_{i=1}, n \geq 3$*, be a sequence of independent Bernoulli trials, each with probability of success* $0 < \pi < 1$*. Let* $\hat{\Pi}_k(\mathbf{X})$ *be the proportion of successes on the subset of trials* $I_k(\mathbf{X})$ *that immediately follow* $k$ consecutive successes, that is, $\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k$ *is a biased estimate of* $\mathbb{\Pi}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv p$ *for all* $k$ *such that* $1 \leq k \leq n - 2$*. In particular,* $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.$

MS note that there does not exists a closed form representation for $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$ for $k > 1$. ($\hat{\Pi}_k(\mathbf{X})$ is the analogue for $\hat{P}^{w|kr}_{MS}$ in my notation for CoD SnD.) Figure 1 shows how the expected value varies as a function of the total number of trials $n$, given $\pi$ and $k > 1$.[^13]

[^13]: I've adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

@@

# library(extrafont)
# library(ggplot2)
# library(latex2exp)
# # set.seed(42)
# # runs <- rerun(
# #   1000,
# #   crossing(
# #     n = 1:50,
# #     k = 2:4,
# #     p = 0.5
# #   ) |>
# #     mutate(
# #       next_p = pmap_dbl(list(n, k, p), ~simulate_post_streak_prob(sims = 1000, n = ..1, k = ..2, p = ..3))
# #     )
# # ) |>
# #   reduce(bind_rows)
# # qs::qsave(runs, 'data/runs_1000x1000_nkp.qs')
# runs <- qs::qread('data/runs_1000x1000_nkp.qs')
# 
# agg_runs <- runs |>
#   filter(!is.nan(next_p)) |>
#   group_by(p, k, n) |>
#   summarize(
#     across(next_p, mean)
#   ) |>
#   ungroup() |>
#   arrange(p, k)
# 
# extrafont::loadfonts(quiet = TRUE)
# theme_set(theme_classic())
# theme_update(
#   text = element_text(family = 'CMU Typewriter Text')
# )
# 
# p <- agg_runs |>
#   arrange(p, k, n) |>
#   mutate(
#     across(k, factor)
#   ) |>
#   ggplot() +
#   aes(x = n, y = next_p, color = k, group = k) +
#   guides(
#     color = guide_legend(
#       title = 'streak length (k)',
#       title.hjust = 0.5,
#       nrow = 1, override.aes = list(size = 3)
#     )
#   ) +
#   geom_line() +
#   geom_hline(aes(yintercept = p)) +
#   labs(x = 'number of trials (n)', y = expression('post-streak probability of success'~(pi))) +
#   theme(
#     legend.position = c(0.5, 0.8)
#   )
# p
# 
# ggsave(
#   p,
#   filename = 'paper/images/pwkr_ms.png',
#   width = 5.5,
#   height = 3
# )
```

```{=tex}
\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials that immediately follow $k$ consecutive successes, $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $n.$}
\end{figure}
```
My context is fundamentally different from that of GVT and MS, both of whom focus on longitudinal data in controlled settings. GVT also perform statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team. The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent.

Consequently, a statistical test of the difference in $\hat{P}^{w|kr}$ and $\hat{P}^{\ell|kr}$, as performed by GVT and MS to evaluate their respective hypothesis regarding $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$ (Theorem 1), is not appropriate.

Let us consider another form of the expected proportion of rounds won immediately after a streak of $k$ round wins in a best-of-$s$ series given the length of the series ($r$ rounds), the "hindsight" proportion $\hat{P}^{w|kr}_0$. The Bernoulli round win probability underlying $\hat{P}^{w|kr}_0$ is

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0^{w|kr} = p_0(\text{win | k, r})=\{
\begin{array}{cl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}.
}
\label{eq:pwkr}
\end{equation}
```
A binomial test for the expected proportion of round wins following streaks of $k$ round wins in a series lasting $r$ rounds, $\hat{P}^{w|kr}_0$, assuming $p^{w|kr}_0$ as the probability of success, can be used to verify that $\hat{P}^{w|kr}_0$ is reasonable as to use as a baseline with which to compare $\hat{P}^{w|kr}_{MS}$. If the binomial test shows that we cannot reject the null hypothesis that the expected $\hat{P}^{w|kr}_0$ is different from the observed proportion $P^{w|kr}$.

A proportion test of $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ can tell us whether the momentum exists in CoD SnD. A null result means we cannot say that there is a difference between the hindsight expected proportion $\hat{P}^{w|kr}_0$ and the expected proportion adjusted for streak-selection bias, $\hat{P}^{w|kr}_0$.

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^14] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^14]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^15] Table \ref{tbl:cod-pw3r-pl3r} shows $P^{w,kr}$ and $P^{l,kr}$ for round win rate and loss rate following $k = 3$ round win streaks respectively.

[^15]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

>>>>

The formula for the expected proportion of series a best-of-$s$ format ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^11] is

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}
```
$s = 11$ rounds for CoD SnD.[^12]

[^12]: $\phi$ and other symbols are selected so as to avoid reserve symbols, like $p$ for usage in other contexts, without risking confusion. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to proportions represent probabilities. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while "bare" symbols mirroring hat-ed symbols convey observational data.

To evaluate the constant round win probability null hypothesis---that is, that the expected and observed round win rates, $\hat{\Phi}(r)$ and $\Phi(r)$ respectively, are equal to one another---we can compute the chi-square goodness of fit statistic, $\chi^2$,

```{=tex}
\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}.
}\label{eq:chi-squ}\end{equation}
```
$r \in R = [6, 7, 8, 9, 10, 11]$ for CoD SnD.

@@

Let us now consider round win rate immediately following a streak of $k$ wins, given that the series lasts $r$ rounds. Removing our knowledge of a streak, we might model the Bernoulli round win probability as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{w|r}_0 = \{
\begin{array}{rcll}
p^{w|r_W} & = & \frac{m}{r}, & \text{team wins series} \\
p^{w|r_L} & = & \frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}
```
using $m$ from Equation \ref{eq:m}. The Bernoulli round loss probability for a series lasting $r$ rounds, $p^{\ell|kr}_0$ (i.e. $\ell = 1 - w, w \in [0, 1]$), can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}_{MS}$, to be strictly less than $p^{w|r}$.[^13][^14]

[^13]: See Appendix E.1 of MS for details on the proof that follows.

[^14]: Here, I choose symbols, e.g. $\pi$, $x$, etc. that mostly do not conflict with syntax in other places. The exceptions are $k$ and $r$, which are used to represent streak length and number of trials globally.

> Theorem 1: *Let* $\mathbf{X} = \{X_i\}^{r}_{i=1}, r \geq 3$*, be a sequence of independent Bernoulli trials, each with probability of success* $0 < \pi < 1$*. Let* $\hat{\Pi}_k(\mathbf{X})$ *be the proportion of successes on the subset of trials* $I_k(\mathbf{X})$ *that immediately follow* $k$ consecutive successes, that is, $\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k$ *is a biased estimate of* $\mathbb{P}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv \pi$ *for all* $k$ *such that* $1 \leq k \leq r - 2$*. In particular,* $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.$

MS notes that there does not exists a closed form representation for $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, henceforth $E^\Pi$, for $k \> 1$. One may recursively build a collection of dictionaries to quantify $E^\Pi$ analytically, or, more reasonably for large values of $n$, one may run simulations to estimate the expected value. Indeed, Figure 1 uses simulation to estimate $E^\Pi$ as a function of the total number of trials $r$, given $\pi$ and $k$.[^15][^16]

[^15]: I've adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

[^16]: Note that $\pi$ and $\hat{\Pi}_k(\mathbf{X})$ are analogues for $p_0^{w|r}$ and $\hat{P}^{w|kr}_{MS}$ in my notation for CoD SnD.

@@

# # set.seed(42)
# # runs <- rerun(
# #   1000,
# #   crossing(
# #     n = 1:50,
# #     k = 2:4,
# #     p = 0.5
# #   ) |>
# #     mutate(
# #       next_p = pmap_dbl(list(n, k, p), ~simulate_post_streak_prob(sims = 1000, n = ..1, k = ..2, p = ..3))
# #     )
# # ) |>
# #   reduce(bind_rows)
# # qs::qsave(runs, 'data/runs_1000x1000_nkp.qs')
# 
# library(ggplot2)
# library(latex2exp)
# runs <- qs::qread('data/runs_1000x1000_nkp.qs')
# 
# agg_runs <- runs |>
#   filter(!is.nan(next_p)) |>
#   group_by(p, k, n) |>
#   summarize(
#     across(next_p, mean)
#   ) |>
#   ungroup() |>
#   arrange(p, k)
# 
# # theme_set(theme_classic())
# 
# p <- agg_runs |>
#   arrange(p, k, n) |>
#   mutate(
#     across(k, factor)
#   ) |>
#   ggplot() +
#   aes(x = n, y = next_p, color = k, group = k) +
#   guides(
#     color = guide_legend(
#       title = 'k',
#       title.hjust = 0.5,
#       nrow = 1, override.aes = list(size = 3)
#     )
#   ) +
#   geom_line() +
#   geom_hline(aes(yintercept = p)) +
#   labs(x = 'r', y = expression(pi)) +
#   theme_classic() +
#   theme(
#     legend.position = c(0.5, 0.8)
#   )
# p
# 
# ggsave(
#   p,
#   filename = 'paper/images/pwkr_ms.png',
#   width = 5.5,
#   height = 3
# )
```

```{=tex}
\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials, $E^\Pi$, that immediately follow $k$ consecutive successes for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $r$.}
\end{figure}
```
My context is fundamentally different from that of GVT and MS, both of whom focus on longitudinal data in controlled settings. GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team. The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{w|kr}$ and $\hat{P}^{\ell|kr}$, as performed by GVT and MS to evaluate their respective hypothesis regarding post-streak success rates is not appropriate.

One might consider another form of the expected proportion of rounds won immediately after a streak of $k$ round wins in a best-of-$s$ series given the length of the series ($r$ rounds), the "notional" proportion $\hat{P}^{w|kr}_0$. The Bernoulli round win probability underlying $\hat{P}^{w|kr}_0$ is

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rcll}
p_0^{w|kr_W} & = & \frac{m - k}{i - k}, & \text{team wins series} \\
p_0^{w|kr_L} & = & \frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}
```
using $m$ from Equation \ref{eq:m}. We can perform a binomial test to test the null hypothesis $H_0 : \omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^17]

[^17]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

If there are $\sigma$ observed successes in a sample of $r$ trials and we expect that there should be $r * \omega_0$, the probability of getting this expected number of successes is

```{=tex}
\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(\sigma) = {\binom {r}{\sigma}} \omega^{\sigma}(1-\omega)^{r-\sigma}.
}\label{eq:binom}\end{equation}
```
Plugging in the notional probability $p^{w|kr}_0$ for the null $\omega_0$ and the observed probability $p^{w|kr} = \frac{N^{w|kr}}{P^{w|kr}}$, where $N^{w|kr}$ is the observed number of instances where a teams wins the round immediately following a $k$-round winning streak in a series lasting $r$ rounds, for the observed $\omega$, we can evaluate whether the observed probability is not significantly different from the notional probability. If we reject this hypothesis, then we might consider team momentum, represented by post-streak success, plausible.

Returning to the streak-selection adjustment framework proposed by MS, one can perform the same binomial test, swapping out $p^{w|kr}_0$ with $p^{w|kr}_{MS}$. Of course, given the caveats mentioned before, the results should be heeded with caution.

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^18] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^18]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^19] Table \ref{tbl:cod-pw3r-pl3r} shows $P^{w,kr}$ and $P^{l,kr}$ for round win rate and loss rate following $k = 3$ round win streaks respectively.

[^19]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.
